---
title: "GEO 200CN Week 1"
author: "Tyler Jackson"
output:
  html_document:
    theme: journal
    toc: true
    toc_float: true
---
# Part 1 

## 1) Read and explore data

There are two text files with data (loblolly.dat) and (ChickWeightSub.csv). 
Write a script (including brief comments) that

### a) Reads them into R
```{r}
#use read.table for the .dat file
loblolly <- read.table("./data/loblolly.dat", header = TRUE)

#does it look good?
head(loblolly) #yes
#read.csv for the csv
chx_wt <- read.csv("./data/ChickWeightSub.csv")

#look good?
head(chx_wt) #yes
```

### b) for both data sets, lists all the variable names
```{r}
#get the variable names using names()
names(loblolly)

names(chx_wt)
```

###c) for each variable in one of the data sets, use a function to compute the min and a function to compute the mean value 

Minimums
```{r}
#calculate the min for each variable in loblolly
min(loblolly$height)
min(loblolly$age)
min(loblolly$Seed)
```

Means
```{r}
#calculate the mean of each variable in loblolly
mean(loblolly$height)
mean(loblolly$age)
mean(loblolly$Seed)
```

###d) now pick one variable from one data set and use functions summary, table, quantile, and boxplot

Summary
```{r}
#summary gives stat data on a variable
summary(chx_wt$Weight)
```

Table
```{r}
#table gives the count for the number of appearances of each value
table(chx_wt$Weight)
```

Quantile
```{r}
#gives quantiles of the values 
quantile(chx_wt$Weight)
```

Boxplot
```{r}
#creates a boxplot with max and min and iqr and median
boxplot(chx_wt$Weight)
```

# Part 2

## 2) Functions

###a) Use the files from lab 1 to compute mean values for each variable using 'apply'

```{r}
#apply, "2" is columnwise
apply(chx_wt, 2, mean)

apply(loblolly, 2, mean)
```


###b) Write a function 'f' that behaves like this:

>f('Jim')

> [1] "hello Jim, how are you"

```{r}

#function asks for a name and paste it into a sentence
f <- function(name) {
  x <- paste0("hello ", name, ", how are you") #paste0 means no spaces
  return(x)
}

f("Jim")
```


###c) Write a function 'sumOfSquares' that behaves like this:

> d <- c(1,5,2,4,6,2,4,5)

> sumOfSquares(d)

> [1] 21.875

```{r}
#subtracts the mean value of all numbers from each number, squares them and sums them
sumOfSquares <- function(n) {
    x <- sum((n - mean(n))^2)
    return(x)
}

#create a vector of numbers
d <- c(1,5,2,4,6,2,4,5)

#pass to the function
sumOfSquares(d)

```

To compute the "sum of squares", subtract the mean value of all numbers from each number. Square these numbers and sum them


## 3) Data preparation. 

Use the data in "students.csv" and "scores.dat" to assess if there is a relation between the "group" or "age" of a student and their average score and for the three subjects (math, english, geography)
```{r}
#read in the data
students <- read.csv("./data/students.csv")

#scores has -999 as a value, replace with NA for later removal from calculations
scores <- read.table("./data/scores.dat", header = TRUE, na.strings = "-999")

head(students)
head(scores)
```

###a) use reshape and merge to combine the two data sets

```{r}
#reshape the data from long to wide
scores_reshape <- reshape(scores, idvar = "student", timevar = "subject", direction = "wide")

#did that work?
head(scores_reshape) #yes

#merge joinkey is students in students and student in the reshape
student_scores <- merge(students, scores_reshape, by.x = "students", by.y = "student")

#check it out
head(student_scores)
```

###b) compute correlation coefficients

```{r}
#minus the first column because it's not a number, complete.obs excludes na
cor(student_scores[,-1], use = "complete.obs")
```

###c) produce a relevant plot

Scatterplot Matrix
```{r}
#create a scatterplot matrix of variables
pairs(group ~ age + score.math + score.english + score.geography, student_scores, na.action = na.omit)
```

Scores by Age
```{r}
library(ggplot2)
library(ggthemes)
#using geom point to plot each subject score by Age
ggplot(student_scores, aes(x = age, y = Score)) +
  geom_point(aes(y = score.math, color = "Math")) +
  geom_point(aes(y = score.geography, color = "Geography")) +
  geom_point(aes(y = score.english, color = "English")) +
  labs(color = "Subject", title = "Subject Scores by Age", subtitle = "Points") +
  theme_gdocs()
```

Points are hard to read, let's try smoothed means
```{r}
#using geom smooth to plot each subject score by Age
ggplot(student_scores, aes(x = age, y = Score)) +
  geom_smooth(aes(y = score.math, color = "Math")) +
  geom_smooth(aes(y = score.geography, color = "Geography")) +
  geom_smooth(aes(y = score.english, color = "English")) +
  labs(color = "Subject", title = "Subject Scores by Age", subtitle = "Smoothed Means") +
  theme_tufte()
```

> Note that the warning message in both refers to the -999 in the original dataset that was coverted to NA on importing the data.
